N-Body problem

### 2014-10-25 article/article.tex
removes sensitivity analysis
jerk en leapfrog

    x(t+ \Delta t) = x(t) + v(t) \Delta t \\
    v(t+ \Delta t) = v(t) + a(t) \Delta t \\

@@

This integrator is appealing for its simplicity, but because it is essentially a first order taylor expansion the expected error is proportional to $\Delta t^2$.

@@

\subsection{Euler}
\label{sec:res:euler}
To verify our model, we have defined the energy metrics. In this experiment the relative energy metric as defined in section~\ref{XXX} is used. In total we have compared four different time-steps to evaluate the performance effect of an increasing/decreasing time-step. The time-steps used are $0.001, 0.01, 0.1, 1$ and figure~\ref{fig:euler} shows a plot of the relative energy error for the \textit{model simulation time}. 
\begin{figure}
    \includegraphics[width=0.9\textwidth]{euler_different_timesteps.png}
    \caption{Relative energy error for the Euler integrator using different time-steps for a fixed total model simulation time.}
    \label{fig:euler}
\end{figure}
As can be seen in figure~\ref{fig:euler} there is a spike around time 18 for time-step scale 1. This error has no apparent reason and in future time-steps it will decrease and stabilize around -1. Furthermore, it is interesting to see that with smaller time-step scales the total relative energy error increases faster. Meaning, that if we run the model for an infinite time the total error will explode and produce no meaning full results. This, while smaller time-step scales should increase the accuracy in terms of relative energy error. Therefore, we can conclude that the increase in accuracy does not outweighs the increase in iterations to complete the calculation.

\subsection{Taylor}
\label{sec:res:taylor}
Using the same energy metrics defined in section~\ref{XX} we can do the same simulation for the Taylor approximation. We expect that adding an extra term in the Taylor expansion would increase our accuracy. And in figure~\ref{fig:taylor} we can see that indeed slower increase in relative energy error than using bigger time-step scales. However, all the time-step scale converge to -1 after about 20\% of the total model simulation time. Also, we noticed that the 0.001 time-step scale converges faster to the -1 relative energy error than the 0.0001 time-step scale. Here, we can observe that decreasing the time-step by 10 does not increase the accuracy enough to compensate for the increase in number of iterations. The could be due to machine-precision, but requires further investigation to be sure.

\begin{figure}
    \includegraphics[width=0.9\textwidth]{jerk_different_timesteps_truncated.png}
    \caption{Relative energy error for the Taylor integrator using different time-steps for a fixed total model simulation time.}
    \label{fig:taylor}
\end{figure}

\subsection{Leapfrog}
\label{sec:res:leapfrog}
Using the Leapfrog integrator, which is a slight modified Euler integrator, we should expect similar results. And as can be seen in figure~\ref{fig:leapfrog}. the relative energy error shows similar results as for the Euler approximation in figure~\ref{fig:euler}. However, the magnitude of the relative energy error is higher with the Leapfrog integrator at around time-step 18. This is probably due to the approximation of the in between time-steps which the Leapfrog integrator uses. Nevertheless, the results are relatively the same as the Euler approximation and does, at least in our simulation, lead to a decrease in relative energy error.

\begin{figure}
    \includegraphics[width=0.9\textwidth]{leapfrog_different_timesteps.png}
    \caption{Relative energy error for the Leapfrog integrator using different time-steps for a fixed total model simulation time.}
    \label{fig:leapfrog}
\end{figure}

\subsection{All models}
\label{sec:res:all}
To compare all the models, i.e. Euler, Taylor and Leapfrog integrator, we setup up an experiment which uses the same time-step for all the models. The time-step is set at $0.01$ and figure~\ref{} shows the relative energy error for all models. Here we can clearly observe that the Taylor integrator gives the best results in terms of relative energy error. And as expected the Leapfrog and Euler integrator are approximately the same. However, the Taylor approximation comes with extra costs in computation time which will be discussed next.

\begin{figure}
    \includegraphics[width=0.9\textwidth]{constant_time.png}
    \caption{Relative energy error for the Euler,Taylor and Leapfrog integrator using fixed time-steps for a fixed total model simulation time.}
    \label{fig:leapfrog}
\end{figure}

\subsection{Complexity}
\label{sec:res:complexity}

>>>>

        x_{t+ 1} = x_t + v_t \Delta t \\
        v_{t+ 1} = v_t + a_t \Delta t \\

@@

This integrator is appealing for its simplicity, but because it is essentially a first order taylor expansion the expected error is proportional to $\Delta t^2$.\\\\
\textbf{Jerk Integrator}\\
It has been shown that through Newton's laws of gravity and mechanics you can calculate a body's acceleration based on the position of the other bodies. Acceleration is the second derivative of position, so it needs to be integrated twice to update the position of a body. \\\\
It is however also possible to directly calculate the third derivative of the position, the so-called jerk. The jerk is also the first derivative of the acceleration and can thus be obtained by differentiating the acceleration once with respect to time.
\begin{align}
    \vec{j_i} = \frac{d}{dt} \vec{a_i} &= \frac{d}{dt} \left( G \sum_{k=1, k \neq i}^n \frac{m_k (\vec{r_k} - \vec{r_i})}{|\vec{r_k} - \vec{r_i}|^3} \right) \\
                                       &= G \sum_{k=1, k \neq i} m_k \left( \frac{\vec{v_k} - \vec{v_i}}{|\vec{r_k} - \vec{r_i}|^3} - 3 \frac{(\vec{r_k} - \vec{r_i})\left((\vec{r_k} - \vec{r_i})\cdot (\vec{v_k} - \vec{v_i}) \right)}{|\vec{r_k} - \vec{r_i}|^5} \right)
\end{align}
Using the calculated jerk, the updated acceleration, velocity and position can be approximated.
        x_{t+ 1} = x_t + v_t \Delta t + \frac{1}{2} a_t (\Delta t)^2 + \frac{1}{6} j_t (\Delta t)^3 \\
        v_{t+ 1} = v_t a_t \Delta t + \frac{1}{2} j_t (\Delta t)^2 \\
        a_{t+ 1} = a_t + j_t \Delta t \\
\\\\
\textbf{Leapfrog Integrator}\\
The leapfrog integrator calculates the positions and velocities of the bodies in a simulated system at interleaved timesteps. 
        x_{t+ 1} = x_t + v_{t+1/2} \Delta t \\
        v_{t+ 1/2} = v_{t-1/2} + a_t \Delta t \\
According to the literature (e.g. \cite{1995ApJ...443L..93H}) this can be rewritten in a form where all variables are evaluated at the same time.
        x_{t+ 1} = x_t + v_t \Delta t  + \frac{1}{2} a_t (\Delta t)^2\\
        v_{t+ 1} = v_t + \frac{a_t + a_{t+1}}{2} \Delta t \\
One of the appealing characteristics of the leapfrog integrator is that it is time-symmetric. This means that if you let it run forward for a certain number of steps and then run it in reverse it should end up at the same initial starting state. This is supposed to improve its energy conservation properties.

@@

\subsection{Euler}
\label{sec:res:euler}
To verify our model, we have defined the energy metrics. In this experiment the relative energy metric as defined in section~\ref{XXX} is used. In total we have compared four different time-steps to evaluate the performance effect of an increasing/decreasing time-step. The time-steps used are $0.001, 0.01, 0.1, 1$ and figure~\ref{fig:euler} shows a plot of the relative energy error for the \textit{model simulation time}. 
\begin{figure}
\label{fig:euler}
    \includegraphics[width=0.9\textwidth]{euler_different_timesteps.png}
    \caption{Relative energy error for the Euler integrator using different time-steps for a fixed total model simulation time.}
\end{figure}

\subsection{Taylor}
\label{sec:res:taylor}

\subsection{Leapfrog}
\label{sec:res:leapfrog}

\subsection{All models}
\label{sec:res:all}
